---
title: "Using `boRing` to improve clustering"
output: rmarkdown::html_vignette
author: 
  - "Simon Garnier"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{Using `boRing` to improve clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates how to use `boRing` during clustering operations to 
determine the "best" number of clusters in a data set. Here, we consider that a
"good" cluster is a cluster that only has one mode. 

```{r setup}
# install.packages("ggplot2")
library(ggplot2)
library(boRing)
```

For this example, we use the `four_Gaussians` data set provided with the 
`boRing` package. This data set was randomly generated from four bivariate 
Gaussians distributions with diffent parameters. It looks like this:

```{r, out.width="75%", fig.align="center", dpi = 300}
data(four_Gaussians)
ggplot(as.data.frame(four_Gaussians)) +
  aes(V1, V2) +
  geom_point(size = 1) +
  coord_equal() +
  theme_bw(base_size = 7)
```

Next, we apply the k-means clustering to this data set with different numbers of
clusters (from 2 to 20). Each time, we measure the boringness index of the 
clusters identified by the algorithm. 

```{r}
k <- 2:20
boring_idx <- data.frame(
  k = rep(k, k),
  idx = NA
)

for (i in seq_along(k)) {
  km <- kmeans(four_Gaussians, k[i], 100, 10)
  boring_idx[boring_idx$k == k[i], 2] <- sapply(
    unique(km$cluster),
    function(j) {
      idx <- km$cluster == j
      boring(four_Gaussians[idx, ])
    }
  )
}
```

Finally, for each number of clusters, we find the cluster with the lowest 
boringness index (that is, the least unimodal-looking cluster of the lot). The 
"best" number of clusters is that for which the lowest boringness index is the 
highest. 

In the graph below, each dot represents the boringness index of each cluster 
identified by the k-means algorithm for diffent number of clusters. The black 
line represents the the cluster with the lowest boringness index in each case. 
The vertical dashed line indicates the estimated "best" number of clusters, four
in this case. 

```{r, out.width="75%", fig.align="center", dpi = 300}
best_k <- k[
  which.max(
    by(boring_idx, boring_idx$k, min)
  )
]

ggplot(boring_idx) +
  aes(k, idx) +
  geom_point(size = 1) +
  geom_vline(xintercept = best_k, linetype = 2) +
  stat_summary(fun = min, geom = "line") +
  coord_cartesian(ylim = c(-1, 1)) +
  labs(x = "Number of clusters", y = "Boringness index") +
  theme_bw(base_size = 7)
```
