---
title: "More `boRing` Examples"
author: 
  - "Jason M. Graham"
  - "Simon Garnier"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{More `boRing` Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette uses the `penguins` data set from the 
[`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/index.html) 
package by [Allison Horst](https://allisonhorst.com/), 
[Alison Hill](https://www.apreshill.com/), and 
[Kristen Gorman](https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php) 
to demonstrate features of `boRing`. 

```{r setup}
# install.packages("palmerpenguins")
# install.packages("ggplot2")
library(palmerpenguins)
library(ggplot2)
library(boRing)
```

The `penguins` data frame contains observations about penguins consisting of 
year, species, sex, name of island where a bird was observed, and four physical
measurements.

```{r, echo=FALSE}
head(penguins) |>
  knitr::kable()
```

Note that there are missing values for some observations. 

If we plot the body mass of penguins versus their flipper length, we can observe
some clustering by species.

```{r, out.width="75%", fig.align="center", dpi = 300}
penguins |>
  ggplot(aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point(size = 1) +
  scale_color_brewer(palette = "Dark2") +
  guides(color = guide_legend(position = "top")) +
  theme_bw(base_size = 7) +
  theme(
    legend.title = element_blank()
  )
```

This data appears to lack unimodality. Let's compute the boringness index for 
the plotted data:

```{r}
penguins |>
  dplyr::select(flipper_length_mm, body_mass_g) |>
  as.matrix() |>
  boring(na_rm = TRUE, conf_level = 0.95)
```

The boringness index value reflects what we observed in the plot. Contrast the
previous result with the boringness index value obtained when we restrict to 
only the Gentoo species:

```{r}
penguins |>
  dplyr::filter(species == "Gentoo") |>
  dplyr::select(flipper_length_mm, body_mass_g) |>
  as.matrix() |>
  boring(na_rm = TRUE, conf_level = 0.95)
```

This boringness index value is consistent with the increased unimodality of the 
data when observing only the single species of Gentoo.

As another illustration, let's examine the relationship between the bill length
and bill depth of the penguins:

```{r, out.width="75%", fig.align="center", dpi = 300}
penguins |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = species)) +
  geom_point(size = 1) +
  scale_color_brewer(palette = "Dark2") +
  guides(color = guide_legend(position = "top")) +
  theme_bw(base_size = 7) +
  theme(
    legend.title = element_blank()
  )
```

Again, this data appears to lacks unimodality but now there is also lower 
density in the center of the cloud of data points. Let's compute the boring 
metric for the plotted data:

```{r}
penguins |>
  dplyr::select(bill_length_mm, bill_depth_mm) |>
  as.matrix() |>
  boring(na_rm = TRUE, conf_level = 0.95)
```

The negative value for the boringness index is consistent with what we observed
in the plot. 

It may be interesting to consider what is happening with regard to the bill
length and bill depth for the Chinstrap species:

```{r}
penguins |>
  dplyr::filter(species == "Chinstrap") |>
  dplyr::select(bill_length_mm, bill_depth_mm) |>
  as.matrix() |>
  boring(na_rm = TRUE, conf_level = 0.95)
```

A relatively low value of unimodality is predicted. This may be explained
looking at the difference in the measurements across sex:

```{r, out.width="75%", fig.align="center", dpi = 300}
penguins |>
  dplyr::filter(species == "Chinstrap") |>
  dplyr::select(bill_length_mm, bill_depth_mm, sex) |>
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = sex)) +
  geom_point(size = 1) +
  scale_color_brewer(palette = "Dark2") +
  guides(color = guide_legend(position = "top")) +
  theme_bw(base_size = 7) +
  theme(
    legend.title = element_blank()
  )
```

Finally, we compute the boringness index for all four physical measurements:

```{r}
penguins |>
  dplyr::select(bill_length_mm, bill_depth_mm, body_mass_g, flipper_length_mm) |>
  as.matrix() |>
  boring(na_rm = TRUE, conf_level = 0.95)
```
