---
title: "How the boringness index is constructed"
output: rmarkdown::html_vignette
author: 
  - "Simon Garnier"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How the boringness index is constructed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how the `boring` function computes the boringness index. 

```{r setup}
# install.packages("ggplot2")
library(ggplot2)
library(MASS)
library(boRing)
```

At the core of the boringness index is the idea that the density of observations
should decrease monotically as one moves away from the center of mass of a
unimodal distribution. Take for instance the following bivariate random Gaussian 
distribution: 

```{r, out.width="75%", fig.align="center", dpi = 300}
mu1 <- c(3, 3)
sigma1 <- matrix(c(3, 0.5, 0.5, 1), 2, 2)
distr1 <- as.data.frame(mvrnorm(500, mu1, sigma1))
com1 <- apply(distr1, 2, mean)

ggplot(distr1) +
  aes(x = V1, y = V2) +
  stat_density_2d(aes(fill = after_stat(level)),
    geom = "polygon",
    n = 100, bins = 10, show.legend = FALSE
  ) +
  geom_point(color = "white", size = 1, alpha = 0.5) +
  geom_point(
    x = com1[1], y = com1[2], shape = 21,
    color = "white", fill = "red", size = 3, stroke = 1
  ) +
  coord_equal(xlim = c(-4, 8), ylim = c(-6, 6)) +
  scale_fill_viridis_c(option = "plasma") +
  theme_bw(base_size = 7)
```

As you can see, the density of points decreases more or less regularly around 
the center of mass of the distribution (the red dot). Let's add a second 
distribution to the mix, different from the first one, and plot the result:  

```{r, out.width="75%", fig.align="center", dpi = 300}
mu2 <- c(-1, -1)
sigma2 <- matrix(c(1, 0.5, 0.5, 3), 2, 2)
distr2 <- rbind(distr1, as.data.frame(mvrnorm(500, mu2, sigma2)))
com2 <- apply(distr2, 2, mean)

ggplot(distr2) +
  aes(x = V1, y = V2) +
  stat_density_2d(aes(fill = after_stat(level)),
    geom = "polygon",
    n = 100, bins = 10, show.legend = FALSE
  ) +
  geom_point(color = "white", size = 1, alpha = 0.5) +
  geom_point(
    x = com2[1], y = com2[2], shape = 21,
    color = "white", fill = "red", size = 3, stroke = 1
  ) +
  coord_equal(xlim = c(-4, 8), ylim = c(-6, 6)) +
  scale_fill_viridis_c(option = "plasma") +
  theme_bw(base_size = 7)
```

Now, as we move away from the center of mass of the distribution, the density of
points first increases and then decreases. `boRing` tries to capture this 
difference in the density pattern to estimate how unimodal (that is, boring) an
empirical distribution is, regardless of its dimensionality. 

The computation of the "boringness" index is done in five simple steps that we 
will demonstrate using the two distributions generated earlier. In the rest of 
the document, we're breaking down the process using commonly available R 
functions. However, under the hood of `boRing` we have reimplemented these 
functions using the [`RcppEigen`](https://github.com/RcppCore/RcppEigen) library
for increased performance. 

---

**Step 1:** `boRing` computes the center of mass and the covariance matrix of
the empirical distribution.

```{r}
com1 <- apply(distr1, 2, mean)
covar1 <- cov(distr1)

com2 <- apply(distr2, 2, mean)
covar2 <- cov(distr2)
```

Note that `boRing` can also use weighted observations if required. Read the help
page of the `boring` function to learn more about this. 

---

**Step 2:** Using the center of mass and the covariance matrix, `boRing`
computes the Mahalanobis distance of each observation to the center of mass of
the distribution. The Mahalanobis distance is the distance of a point to the
center of mass divided by the width of the ellipsoid (which axes are determined
by the covariance matrix) in the direction of the point.

```{r}
# The `mahalanobis` function returns the squared Mahalanobis distance, hence the
# square root transform below.
mahalanobis_dist1 <- sqrt(mahalanobis(distr1, com1, covar1))
mahalanobis_dist2 <- sqrt(mahalanobis(distr2, com2, covar2))
```

---

**Step 3:** `boRing` reorders the observations based on their Mahalanobis 
distance (from closest to the center of mass to furthest away).

```{r}
mahalanobis_dist1 <- sort(mahalanobis_dist1)
mahalanobis_dist2 <- sort(mahalanobis_dist2)
```

--- 

**Step 4:** `boRing` computes the density of observations for ellipsoids of 
different sizes around the center of mass.

```{r, out.width="75%", fig.align="center", dpi = 300}
# First, we compute the volume of the unit ellipsoids corresponding to the
# covariance matrices obtained earlier.
r <- 1

dim1 <- nrow(covar1)
v_unit1 <- (r^dim1 * pi^(dim1 / 2) / gamma(dim1 / 2 + 1)) / sqrt(det(covar1))

dim2 <- nrow(covar2)
v_unit2 <- (r^dim2 * pi^(dim2 / 2) / gamma(dim2 / 2 + 1)) / sqrt(det(covar2))

# Then, we compute the volume of ellipsoids with radii corresponding to the
# Mahalanobis distances of each observations.
ve1 <- v_unit1 * mahalanobis_dist1^dim1
ve2 <- v_unit2 * mahalanobis_dist2^dim2

# Finally, we compute the density of points within each of these ellipsoids...
dens1 <- 1:nrow(distr1) / ve1
dens2 <- 1:nrow(distr2) / ve2

# ... and plot it against the Mahalanobis distances.
dt <- data.frame(
  mahalanobis = c(mahalanobis_dist1, mahalanobis_dist2),
  density = c(dens1, dens2),
  type = c(rep("Unimodal", length(dens1)), rep("Bimodal", length(dens2)))
)

ggplot(dt) +
  aes(x = mahalanobis, y = density, color = type) +
  geom_point(size = 1) +
  labs(
    x = "Mahalanobis distance",
    y = "Density of observations \u2264 Mahalanobis distance"
  ) +
  scale_color_brewer(palette = "Dark2") +
  guides(color = guide_legend(position = "inside")) +
  theme_bw(base_size = 7) +
  theme(
    legend.justification = c(1, 1),
    legend.position.inside = c(0.99, 0.99),
    legend.title = element_blank()
  )
```

As you can clearly see (hopefully), the density of observations decreases more
or less monotonically with the Mahalanobis distance in the case of the first
empirical distribution (drawn from a known unimodal distribution). In the case
of the second empirical distribution (drawn from two different unimodal 
distributions), however, it first increases and then decreases.

--- 

**Step 5:** `boRing` computes the boringness index as the Spearman correlation
between the (negative) Mahalanobis distances and the density of observations
within the ellipsoid corresponding to each of these distances.

```{r}
boring1 <- cor(-mahalanobis_dist1, dens1, method = "spearman")
boring1

boring2 <- cor(-mahalanobis_dist2, dens2, method = "spearman")
boring2
```

The Spearman correlation is rank-based and, therefore, estimates the strength
and direction of monotonic association between two variables. Therefore, under
our starting assumption that the density of observations should decrease
monotically as one moves away from the center of mass of a unimodal 
distribution, higher values of the correlation above will, mechanically,
correspond to empirical distributions with higher degree of unimodality. 

---